---
title: "Exploratory analysis of a public 16S rRNA microbiome dataset"
author: "Umutcan Vural"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
    number_sections: true
geometry: margin=1in
---

## Readme & Timeline

This project conducts an exploratory downstream analysis of processed 16S rRNA amplicon sequencing data from the GlobalPatterns dataset. It evaluates sequencing depth, alpha diversity, beta diversity, taxonomic composition, and differences in microbial community structure across sample types using R.

## Settings

```{r setup-packages, message=FALSE, warning=FALSE}
required_packages <- c(
  "phyloseq",
  "vegan",
  "ape",
  "janitor",
  "here",
  "microbiome",
  "ggplot2",
  "dplyr",
  "tibble",
  "tidyr",
  "readr",
  "stringr",
  "forcats",
  "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Missing packages: ",
    paste(missing_packages, collapse = ", "),
    ". Install dependencies as described in README.md before rendering the report."
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))
```

## Required packages

**Installation:** See scripts/install_dependencies.R

- **ggplot2**, **dplyr**, **readr**, **tidyr**, **tibble**, **stringr**, and **forcats**: provide data manipulation, import, reshaping, and visualisation tools.

- **rmarkdown** and **knitr**: render the reproducible HTML report. `tinytex` is optional and is required only for PDF output.

- **phyloseq**: Integrates microbiome abundance data, taxonomy, phylogenetic information, and sample metadata into a single object for ecological analysis.

- **vegan**: Provides ecological statistics, including Bray–Curtis dissimilarity calculations, ordination support, and PERMANOVA.

- **microbiome**: Provides microbiome-specific functions for abundance transformation, prevalence filtering, and community composition analysis.

- **ape**: Provides functions for reading, handling, and analysing phylogenetic trees when a tree is included in the dataset.

- **janitor**: Provides convenient functions for cleaning and standardising variable and column names.

- **here**: Creates reliable relative file paths within the project folder, improving reproducibility across computers.

# Introduction and Research Question

**Research question:**

> How do microbial alpha diversity, beta diversity, and taxonomic composition vary across environmental and host-associated sample types in the GlobalPatterns dataset?

Microbial communities differ substantially depending on the environment or host they are associated with. This analysis explores whether samples from different environmental and host-associated sources (e.g., soil, ocean, feces, skin) differ in:

- **Alpha diversity** — diversity *within* a sample (richness/evenness)
- **Beta diversity** — diversity *between* samples (community structure)
- **Taxonomic composition** — which taxa dominate each sample type

# Dataset Description

`GlobalPatterns` is a widely used example 16S rRNA amplicon dataset distributed with the `phyloseq` package (Caporaso et al. 2011). It contains high-throughput sequencing data from 26 samples spanning diverse environmental and host-associated sources, along with an associated phylogenetic tree and taxonomy assignments.


# Workflow Setup

```{r load-workflow-outputs}
alpha_diversity <- readr::read_csv(
  "results/alpha_diversity.csv",
  show_col_types = FALSE
)

pcoa_coordinates <- readr::read_csv(
  "results/pcoa_coordinates.csv",
  show_col_types = FALSE
)

var_pc1 <- pcoa_coordinates$VarExplainedPCoA1[1]
var_pc2 <- pcoa_coordinates$VarExplainedPCoA2[1]
```

```{r validate-workflow-outputs}
stopifnot(
  nrow(alpha_diversity) == nrow(pcoa_coordinates),
  all(c("SampleID", "Observed", "Shannon", "Simpson", "SampleType") %in%
        names(alpha_diversity)),
  all(c("SampleID", "PCoA1", "PCoA2", "SampleType") %in%
        names(pcoa_coordinates))
)
```

# Load GlobalPatterns

```{r load-data}
data("GlobalPatterns", package = "phyloseq")
GlobalPatterns
```

```{r load-cleaned-workflow-data}
ps_clean <- readRDS("results/cleaned_data.rds")
ps_clean
```

# Inspect Data Structure

```{r structure}
ntaxa(GlobalPatterns)
nsamples(GlobalPatterns)
rank_names(GlobalPatterns)

# Zero-count taxa (candidates for filtering later)
sum(taxa_sums(GlobalPatterns) == 0)

# Any missing values in the OTU table?
any(is.na(otu_table(GlobalPatterns)))
```

# Inspect Sample Metadata

```{r metadata}
sample_data(GlobalPatterns) %>%
  as("data.frame") %>%
  glimpse()

# Sample types available for grouping in later analyses
sample_data(GlobalPatterns)$SampleType %>% unique()

# Sample count per SampleType
sample_data(GlobalPatterns) %>%
  as("data.frame") %>%
  count(SampleType, sort = TRUE)
```

# Sequencing-Depth Assessment

Sequencing depth (library size = total reads per sample) varies naturally across samples and sample types. Low-depth samples can artificially deflate alpha diversity estimates and distort beta diversity distances, so we inspect depth here before deciding on any filtering or exclusion.

```{r seq-depth}
# Build a depth table: one row per sample, with SampleType attached
depth_df <- tibble(
  SampleID   = sample_names(GlobalPatterns),
  LibSize    = sample_sums(GlobalPatterns),
  SampleType = sample_data(GlobalPatterns)$SampleType
)

depth_df %>% arrange(LibSize)

```

```{r seq-depth-summary}
# Numeric summary overall
summary(depth_df$LibSize)

# Numeric summary by SampleType
depth_df %>%
  group_by(SampleType) %>%
  summarise(
    n      = n(),
    min    = min(LibSize),
    median = median(LibSize),
    mean   = mean(LibSize),
    max    = max(LibSize),
    .groups = "drop"
  ) %>%
  arrange(median)
```

```{r seq-depth-hist}
ggplot(depth_df, aes(x = LibSize)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Sequencing Depth Across All Samples",
    x = "Library size (total reads)",
    y = "Number of samples"
  ) +
  theme_minimal()
```

```{r seq-depth-boxplot}
ggplot(depth_df, aes(x = reorder(SampleType, LibSize, median), y = LibSize)) +
  geom_boxplot(outlier.shape = NA, fill = "grey85") +
  geom_jitter(aes(color = SampleType), width = 0.15, size = 2, show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Sequencing Depth by Sample Type",
    x = "Sample type",
    y = "Library size (total reads)"
  ) +
  theme_minimal()
```

```{r seq-depth-outliers}
# Flag samples below the 10th percentile of overall library size as a
# starting heuristic; adjust threshold after visual inspection above.
low_depth_threshold <- quantile(depth_df$LibSize, 0.10)

depth_df %>%
  filter(LibSize < low_depth_threshold) %>%
  arrange(LibSize)
```

```{r seq-depth-ratio-gaps}
# Check for natural gaps / isolated low-depth failures by looking at the
# ratio between each sample and the next-lowest one, once sorted by depth.
depth_df %>%
  arrange(LibSize) %>%
  mutate(
    PreviousDepth   = lag(LibSize),
    RatioToPrevious = LibSize / PreviousDepth
  )
```

```{r seq-depth-rarecurve, fig.width=8, fig.height=6}
# Rarefaction curves: if a sample's curve plateaus well before its total
# depth, its diversity estimate is not depth-limited. If curves for
# low-depth samples end while still rising steeply, that IS a depth-driven
# failure signal (unlike a simple ratio/threshold check).
otu_mat <- as(otu_table(GlobalPatterns), "matrix")
if (taxa_are_rows(GlobalPatterns)) otu_mat <- t(otu_mat)

rarecurve(
  otu_mat,
  step      = 1000,
  label     = TRUE,
  cex       = 0.6,
  col       = "steelblue",
  main      = "Rarefaction Curves by Sample"
)
abline(v = min(depth_df$LibSize), col = "red", lty = 2)
```

```{r seq-depth-rarecurve-zoom, fig.width=8, fig.height=6}
# Zoom in on the lowest-depth samples only, so the slope near each curve's
# endpoint can actually be judged (the full 26-sample plot is too compressed
# to read this reliably).
low_depth_samples <- depth_df %>%
  arrange(LibSize) %>%
  slice_head(n = 6) %>%
  pull(SampleID)

rarecurve(
  otu_mat[low_depth_samples, ],
  step  = 500,
  label = TRUE,
  cex   = 0.8,
  col   = "steelblue",
  main  = "Rarefaction Curves: 6 Lowest-Depth Samples"
)
```

**Observed result:** The dashed red line marks the minimum observed library size (58,688 reads; `TRRsed1`). Curves for the lowest-depth samples were still increasing near their endpoints, suggesting that sequencing did not fully capture all rare features. Alpha-diversity estimates should therefore be interpreted as conservative lower-bound estimates of richness.

# Filtering Decision

### Decision on low-depth samples

All 26 samples were retained. Although three samples fell below the 10th percentile of sequencing depth, none showed a clear isolated low-depth failure, and excluding samples would further reduce already small group sizes. Alpha diversity was calculated from rarefied counts at the minimum observed depth, while beta diversity used relative-abundance profiles. Rarefaction curves for the lowest-depth samples were still increasing at their endpoints; richness estimates should therefore be interpreted as conservative lower bounds. Features with zero counts across all samples were removed before downstream analysis.

### Taxon-level filtering

Separately from sample filtering, taxa with zero counts across all samples add no information and are removed before downstream analysis.

```{r filtering-taxa}
# Remove taxa with zero total counts across all samples
GlobalPatterns_filt <- prune_taxa(taxa_sums(GlobalPatterns) > 0, GlobalPatterns)

ntaxa(GlobalPatterns)       # before
ntaxa(GlobalPatterns_filt)  # after

before <- ntaxa(GlobalPatterns)
after  <- ntaxa(GlobalPatterns_filt)

percent_removed <- (before - after) / before * 100
percent_retained <- after / before * 100
percent_removed
percent_retained

```

# Alpha diversity

> How diverse is each individual microbiome sample?

Alpha diversity describes the diversity within an individual microbiome sample. This analysis uses rarefied feature counts, in which all samples contain the same number of reads, to reduce the influence of unequal sequencing depth.

Two metrics are examined:

- **Observed richness**: the number of observed microbial features in a sample.
- **Shannon diversity**: a measure that reflects both the number of features and the evenness of their relative abundances.

```{r alpha-diversity-workflow-data}
# Alpha-diversity metrics were generated by the Snakemake workflow
# and loaded from results/alpha_diversity.csv near the start of this report.

alpha_diversity
```

```{r observed-richness-plot, fig.width=10, fig.height=6}
ggplot(
  alpha_diversity,
  aes(x = SampleType, y = Observed, fill = SampleType)
) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2.5, alpha = 0.85) +
  labs(
    title = "Observed microbial richness by sample type",
    subtitle = "Calculated from rarefied 16S feature counts",
    x = "Sample type",
    y = "Observed features"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r shannon-diversity-plot, fig.width=10, fig.height=6}
ggplot(
  alpha_diversity,
  aes(x = SampleType, y = Shannon, fill = SampleType)
) +
  geom_boxplot(show.legend = FALSE, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2.5, alpha = 0.85) +
  labs(
    title = "Shannon diversity by sample type",
    subtitle = "Calculated from rarefied 16S feature counts",
    x = "Sample type",
    y = "Shannon diversity"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r alpha-diversity-summary}
alpha_summary <- alpha_diversity |>
  group_by(SampleType) |>
  summarise(
    Samples = n(),
    MedianObserved = median(Observed),
    MedianShannon = median(Shannon),
    MedianSimpson = median(Simpson),
    .groups = "drop"
  ) |>
  arrange(desc(MedianShannon))

alpha_summary
```

**Observed result:** Soil samples showed the highest richness and Shannon diversity, indicating diverse and relatively even communities. Freshwater creek samples had high richness but low Shannon diversity, suggesting uneven feature abundances. Results are descriptive because groups were small and heterogeneous.

### Statistical test

Kruskal–Wallis tests assessed overall differences across sample types because group sizes were small and unequal (n = 2–4) and normality could not be assumed.

```{r alpha-kruskal}
kruskal.test(Observed ~ SampleType, data = alpha_diversity)
kruskal.test(Shannon  ~ SampleType, data = alpha_diversity)
kruskal.test(Simpson  ~ SampleType, data = alpha_diversity)
```

**Result:** Alpha diversity differed across sample types for all three metrics: Observed richness (\(\chi^2 = 21.33\), \(df = 8\), \(p = 0.0063\)), Shannon diversity (\(\chi^2 = 22.13\), \(df = 8\), \(p = 0.0047\)), and Simpson diversity (\(\chi^2 = 21.42\), \(df = 8\), \(p = 0.0061\)). This supports the descriptive pattern of higher diversity in soil and lower diversity in some host-associated or mock samples.

The significant omnibus tests indicate that at least one sample type differs from another, but do not identify specific group pairs. Because groups contain only 2–4 samples, results are exploratory and no pairwise post-hoc comparisons were conducted.

# Beta diversity / PCoA

Beta diversity assesses differences in microbial community composition between samples. Bray–Curtis dissimilarity was calculated from relative-abundance profiles to account for unequal sequencing depth while retaining all observed taxa. PCoA was used to visualise the resulting between-sample relationships.

```{r pcoa-workflow-data}
# PCoA coordinates were generated by the Snakemake workflow from
# Bray-Curtis distances calculated on relative-abundance profiles.
# `pcoa_coordinates`, `var_pc1`, and `var_pc2` were loaded near the
# beginning of this report.

pcoa_coordinates
```

```{r beta-pcoa-plot, fig.width=8, fig.height=6}
ggplot(
  pcoa_coordinates,
  aes(x = PCoA1, y = PCoA2, colour = SampleType)
) +
  geom_point(size = 3) +
  labs(
    title = "PCoA of Bray–Curtis dissimilarity",
    x = paste0("PCoA 1 (", var_pc1, "%)"),
    y = paste0("PCoA 2 (", var_pc2, "%)")
  ) +
  theme_minimal()
```

```{r beta-diversity-statistics}
# Beta-diversity statistics use the workflow-validated phyloseq object.
GlobalPatterns_relabund <- transform_sample_counts(
  ps_clean,
  function(x) x / sum(x)
)

bray_dist <- phyloseq::distance(
  GlobalPatterns_relabund,
  method = "bray"
)

# Convert metadata to a base data.frame and align it to the distance matrix.
sample_df <- as.data.frame(sample_data(ps_clean))
sample_df <- sample_df[labels(bray_dist), , drop = FALSE]
```

**Observed result:** Samples positioned closer together in the PCoA plot have more similar microbial compositions, while more distant samples have more dissimilar relative-abundance profiles.

**Result:** The first two PCoA axes explained `r var_pc1`% and `r var_pc2`% of Bray–Curtis compositional variation, respectively (`r round(var_pc1 + var_pc2, 1)`% combined). The two-dimensional plot provides a useful visual summary, although most community variation is distributed across additional dimensions.

# PERMANOVA

PCoA visualises compositional relationships but does not test whether differences among sample types are statistically significant. PERMANOVA (`adonis2()`) tests whether microbial community composition is associated with `SampleType` by comparing within-group and between-group Bray–Curtis dissimilarities using permutations.

```{r permanova-test}
set.seed(123)

stopifnot(
  is.data.frame(sample_df),
  identical(rownames(sample_df), labels(bray_dist)),
  "SampleType" %in% names(sample_df)
)

permanova_result <- vegan::adonis2(
  bray_dist ~ SampleType,
  data = sample_df,
  permutations = 999
)

permanova_result
```

**Observed result::** the `R2` column shows the proportion of total community-composition variance explained by `SampleType`; the `p-value` column tests whether this is greater than expected by chance, based on 999 random permutations of group labels.

### Checking the PERMANOVA assumption

PERMANOVA assumes similar within-group dispersion (spread) across groups; if dispersion differs substantially, a significant PERMANOVA result could reflect differences in spread rather than differences in group centroids (mean composition). This is checked with `betadisper()`.

```{r permanova-dispersion}
dispersion <- betadisper(bray_dist, sample_df$SampleType)
anova(dispersion)
```

PERMANOVA indicated that microbial community composition was strongly associated with `SampleType` (\(R^2 = 0.731\), \(F = 5.78\), \(p = 0.001\)). However, multivariate dispersion also differed among sample types (\(F = 5.81\), \(p = 0.0012\)), indicating unequal within-group variability.

Therefore, the PERMANOVA result should be interpreted cautiously: observed separation may reflect both differences in average community composition and differences in within-group heterogeneity. Some sample types formed tight clusters of compositionally similar replicates, whereas others showed greater internal variation. Given the small and unequal group sizes (\(n = 2\)–\(4\)), results are exploratory rather than confirmatory.

```{r permanova-dispersion-plot, fig.width=9, fig.height=5}
dispersion_df <- tibble(
  SampleID   = names(dispersion$distances),
  Distance   = dispersion$distances,
  SampleType = sample_df$SampleType
)

ggplot(dispersion_df, aes(x = reorder(SampleType, Distance, median), y = Distance)) +
  geom_boxplot(fill = "grey85", outlier.shape = NA) +
  geom_jitter(aes(color = SampleType), width = 0.15, size = 2, show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Within-Group Dispersion by Sample Type",
    subtitle = "Distance to group centroid (Bray-Curtis)",
    x = "Sample type",
    y = "Distance to centroid"
  ) +
  theme_minimal()
```

This plot identifies which groups are driving the significant dispersion result. With only 2–4 points per group, the boxplots themselves are hard to read reliably — a numeric summary is more informative here.

**Observed Result:** Small distance → samples within that group are similar to each other → tight group / low within-group heterogeneity Large distance → samples within that group vary more from one another → spread-out group / high within-group heterogeneity

```{r permanova-dispersion-plot-v2, fig.width=9, fig.height=5}
dispersion_df <- data.frame(
  SampleType = sample_df$SampleType,
  DistanceToCentroid = dispersion$distances
)

ggplot(
  dispersion_df,
  aes(x = SampleType, y = DistanceToCentroid, colour = SampleType)
) +
  geom_jitter(width = 0.12, size = 3, alpha = 0.85, show.legend = FALSE) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4,
    colour = "black"
  ) +
  labs(
    title = "Within-sample-type dispersion of microbial communities",
    subtitle = "Bray–Curtis distance from each sample to its sample-type centroid",
    x = "Sample type",
    y = "Distance to group centroid"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r permanova-dispersion-median}
dispersion_df %>%
  group_by(SampleType) %>%
  summarise(
    n              = n(),
    MedianDistance = median(DistanceToCentroid),
    MinDistance    = min(DistanceToCentroid),
    MaxDistance    = max(DistanceToCentroid),
    Range          = MaxDistance - MinDistance,
    .groups = "drop"
  ) %>%
  arrange(desc(MedianDistance))
```

The table summarises each sample type’s distance to its group centroid, providing a clearer view of within-group dispersion than boxplots with few observations.

`Freshwater` and `Tongue` each contain two samples. Their near-zero distance range is a geometric property of two-sample groups, where both points are equidistant from their midpoint, rather than evidence of low biological variability.

Among groups with at least three samples, Mock had the lowest median distance to centroid (0.040), followed by Freshwater (creek; 0.108), indicating relatively consistent composition within these groups. Feces had the highest median distance (0.439), while Skin showed the widest distance range (0.226–0.598), indicating greater within-group heterogeneity. These descriptive patterns are consistent with the significant dispersion test but should be interpreted cautiously because group sizes remain small.

# Taxonomic composition

Feature-level relative abundances were aggregated to the phylum level and visualised as stacked bar charts. The ten most abundant phyla were shown individually, with all remaining phyla grouped as “Other.”

```{r taxonomic-composition-prep}
# Use the workflow-validated object and convert counts to relative abundance.
ps_relative_taxonomy <- transform_sample_counts(
  ps_clean,
  function(x) x / sum(x)
)

# Agglomerate to phylum level
ps_phylum <- tax_glom(GlobalPatterns_relabund, taxrank = "Phylum")

tax_df <- as.data.frame(tax_table(ps_phylum)) %>%
  rownames_to_column("OTU")

# Determine the 10 most abundant phyla by mean relative abundance.
# IMPORTANT: taxa_sums() names are OTU IDs, not phylum names, so they
# must be resolved via tax_df before comparing against the Phylum column.
phylum_means <- taxa_sums(ps_phylum) / nsamples(ps_phylum)
top10_otu    <- names(sort(phylum_means, decreasing = TRUE))[1:10]
top10_phyla  <- tax_df$Phylum[match(top10_otu, tax_df$OTU)] %>% as.character()

top10_phyla  # sanity check: should be 10 readable phylum names, e.g. "p__Bacteroidetes"

# Pull out taxonomy + abundance as a long data frame for plotting, since
# this gives explicit control over the "Other" grouping
otu_df <- as.data.frame(otu_table(ps_phylum))
if (!taxa_are_rows(ps_phylum)) otu_df <- t(otu_df)
otu_df <- otu_df %>%
  as.data.frame() %>%
  rownames_to_column("OTU") %>%
  pivot_longer(-OTU, names_to = "SampleID", values_to = "RelAbundance")

phylum_long <- otu_df %>%
  left_join(tax_df %>% select(OTU, Phylum), by = "OTU") %>%
  left_join(
    sample_data(ps_phylum) %>% as("data.frame") %>%
      rownames_to_column("SampleID") %>%
      select(SampleID, SampleType),
    by = "SampleID"
  ) %>%
  mutate(Phylum = ifelse(Phylum %in% top10_phyla, as.character(Phylum), "Other")) %>%
  group_by(SampleID, SampleType, Phylum) %>%
  summarise(RelAbundance = sum(RelAbundance), .groups = "drop")

# Sanity check: should show 11 distinct values (10 phyla + "Other"),
# not a single dominant category
phylum_long %>% count(Phylum, sort = TRUE)
```

```{r taxonomic-composition-plot, fig.width=11, fig.height=7}
ggplot(phylum_long, aes(x = SampleID, y = RelAbundance, fill = Phylum)) +
  geom_col() +
  facet_wrap(~ SampleType, scales = "free_x", nrow = 2) +
  scale_fill_brewer(palette = "Set3") +
  theme_bw() +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title    = "Taxonomic composition at phylum level",
    subtitle = "Top 10 phyla by mean relative abundance; remaining phyla grouped as \"Other\"",
    x = "Sample",
    y = "Relative abundance",
    fill = "Phylum"
  )
```

### Dominant phyla by sample type

```{r taxonomic-composition-summary}
phylum_long %>%
  group_by(SampleType, Phylum) %>%
  summarise(MeanRelAbundance = mean(RelAbundance), .groups = "drop") %>%
  group_by(SampleType) %>%
  slice_max(MeanRelAbundance, n = 3) %>%
  arrange(SampleType, desc(MeanRelAbundance))
```

The table summarises the three most abundant phyla within each sample type.

**Result:** Feces samples were dominated by Firmicutes (54.8%) and Bacteroidetes (35.2%). Freshwater creek samples were dominated by Cyanobacteria (62.6%), followed by Proteobacteria (18.6%). Non-creek freshwater samples were more balanced, with Actinobacteria (37.1%), Cyanobacteria (30.2%), and Proteobacteria (16.7%) as the most abundant phyla. Mock communities were dominated by a small set of phyla, including Firmicutes (41.4%). These patterns help explain the observed diversity results: Cyanobacteria dominance is consistent with the high richness but lower Shannon diversity of freshwater creek samples, while the consistent compositions of creek and mock samples align with their lower within-group dispersion.

# Interpretation and limitations

This analysis applied a standard downstream 16S microbiome workflow: sequencing-depth assessment, conservative filtering, alpha diversity, beta diversity, and phylum-level taxonomic composition.

Overall, sample type was associated with both alpha diversity and community composition. Soil samples showed high richness and evenness, whereas freshwater creek samples combined high richness with uneven taxonomic composition dominated by Cyanobacteria. Sample type was strongly associated with Bray–Curtis community composition, although unequal within-group dispersion means this result should be interpreted cautiously.

## Limitations

- The analysis used preprocessed 16S data and did not reproduce raw-read processing, denoising, chimera removal, or taxonomy assignment.
- Sequencing counts are compositional; relative abundance does not measure absolute microbial abundance.
- The dataset includes heterogeneous environments and small, unequal sample-type groups, so findings are descriptive and exploratory.
- Alpha diversity was calculated from rarefied counts. Rarefaction standardises depth but discards reads, and low-depth samples may underestimate true richness.
- Dispersion summaries for Freshwater and Tongue are not informative because each group contains only two samples.
- No differential-abundance testing was performed.

# Reproducibility information

This report was generated in R. The session information below records the R version, operating system, and package versions used for the analysis.

```{r session-info}
sessionInfo()
```
